---
title: "Predator and STR Clones monoculture growth workflow iii) analysis of growth summary data"
author: "Shane Hogle"
date: today
---

This notebook plots and performs some analysis on the growthrate and AUC estimates from the prior notebook

# Setup

Load required packages

```{r}
library(here)
library(tidyverse)
library(stringr)
library(stringi)
library(lubridate)
library(fs)
source(here::here("R", "utils_generic.R"))

data_raw <- here::here("_data_raw", "monocultures", "20230112_clone_phenotyping")
```

# Read data

Experiment information (e.g., which wells correspond to which clones from which condition)

```{r}
#| output: false
#| warning: false
#| error: false
mddf <- read_tsv(
  here::here(data_raw, "metadata.tsv"),
  col_types = list(
    plate = col_character(),
    well = col_character(),
    clone_id = col_double(),
    experiment_treatment = col_character(),
    LTCSE_treatment = col_character(),
    replicate = col_character(),
    lagtime = col_time(format = "%T"),
    max_growth_death_rate = col_double(),
    time_to_stationary = col_time(format = "%T")
  )
) %>%
  mutate(
    measure_env_short = case_when(
      measure_env == "bact" ~ "Meas: B",
      measure_env == "bact_pred" ~ "Meas: BP",
      measure_env == "bact_strep" ~ "Meas: BS",
      measure_env == "bact_pred_strep" ~ "Meas: BPS"
    )
  ) %>%
  mutate(
    home_env_short = case_when(
      evolution_env == "anc" ~ "Home: Anc",
      evolution_env == "bact" ~ "Home: B",
      evolution_env == "bact_pred" ~ "Home: BP",
      evolution_env == "bact_strep" ~ "Home: BS",
      evolution_env == "bact_pred_strep" ~ "Home: BPS"
    )
  ) %>%
  mutate(
    measure_env_short = factor(
      measure_env_short,
      levels = c("Meas: B", "Meas: BP", "Meas: BS", "Meas: BPS")
    ),
    home_env_short = factor(
      home_env_short,
      levels = c("Home: Anc", "Home: B", "Home: BP", "Home: BS", "Home: BPS")
    ),
    replicate = factor(replicate),
    clone_id = factor(clone_id)
  )
```

Blacklist (i.e., samples that have been manually inspected and should be excluded due to strange features)

```{r}
#| output: false
#| warning: false
#| error: false
blacklist <- read_tsv(here::here(data_stp_cl, "blacklist.tsv"))
```

Growth summary data 

```{r}
#| output: false
#| warning: false
#| error: false
auc_grates <- readr::read_tsv(here::here(data_stp_cl, "auc_mumax_results.tsv"))
```

# Formatting

Anti join black list to growth curves to remove the bad growth curves and join to experiment metadata

```{r}
auc_gr_md <- left_join(auc_grates, mddf, by = join_by(plate, well)) %>%
  anti_join(blacklist, by = join_by(plate, well))
```

Mean center the AUC and growth rates to the mean of all clones from all evolutionary histories nested by measurement environment

```{r}
auc_gr_md_subset <- auc_gr_md %>%
  dplyr::select(
    auc,
    mumax,
    measure_env,
    evolution_env,
    measure_env_short,
    home_env_short,
    replicate,
    clone_id
  )

auc_gr_md_subset_summary <- auc_gr_md_subset %>%
  filter(measure_env == evolution_env) %>%
  summarize(
    auc_mn = mean(auc),
    auc_sd = sd(auc),
    mu_mn = mean(mumax),
    mu_sd = sd(mumax),
    .by = c(measure_env)
  )

auc_gr_md_subset_summary_z <- left_join(auc_gr_md_subset, auc_gr_md_subset_summary, by = join_by(measure_env)) %>%
  mutate(z_mu = (mumax - mu_mn), #/ mu_sd
         z_auc = (auc - auc_mn)) %>% #/auc_sd
  pivot_longer(cols = starts_with("z"),
               names_to = "metric",
               values_to = "z")
```

# Plot

Function to plot metric of choice

```{r}
cloneplot <- function(df, metric, ylab){
  df %>% 
    filter(metric == {{ metric }}) %>% 
    ggplot2::ggplot() +
    ggplot2::geom_hline(yintercept = 0) +
    ggplot2::geom_boxplot(aes(y = z, x = home_env_short), color = "red", outliers = FALSE, position = position_dodge(width = 0.5)) +
    ggplot2::geom_jitter(aes(y = z, x = home_env_short, color = replicate), width = 0.25, height = 0, alpha = 0.25) +
    ggplot2::labs(x = "Home environment", y = ylab, fill="") + 
    ggplot2::facet_grid(measure_env_short ~ ., scales = "free_y") +
    ggplot2::theme_bw()
}

pclone_auc <- cloneplot(auc_gr_md_subset_summary_z, "z_auc", "Meaurement env mean centered AUC")
pclone_mu <- cloneplot(auc_gr_md_subset_summary_z, "z_mu", "Meaurement env mean centered μ")
```

::: {#fig-01}
```{r}
#| fig.width: 7
#| fig.height: 8
#| echo: false
#| warning: false
pclone_auc
```
Mean centered AUC (area under the growth curve) for 32 randomly picked clones from each replicate (A-F) of each home YSK environment and measurement condition pair. Points are individual measurements (colored by replicate) while boxplots represent the mean and 1-3 quartiles with whiskers representing 1.5 $\times$ the interquartile range. Heavy black line is the Measurement Environment mean.
:::


::: {#fig-02}
```{r}
#| fig.width: 7
#| fig.height: 8
#| echo: false
#| warning: false
pclone_mu
```
Mean centered maximum specific growth rate (μ max) for 32 randomly picked clones from each replicate (A-F) of each home YSK environment and measurement condition pair. Points are individual measurements (colored by replicate) while boxplots represent the mean and 1-3 quartiles with whiskers representing 1.5 $\times$ the interquartile range. Heavy black line is the Measurement Environment mean.
:::


```{r}
ggplot2::ggsave(
  here::here(figs_stp, "monoculture_clone_phenotyping_auc.svg"),
  pclone_auc,
  width = 7,
  height = 8,
  units = "in",
  device = "svg"
)

ggplot2::ggsave(
  here::here(figs_stp, "monoculture_clone_phenotyping_auc.png"),
  pclone_auc,
  width = 7,
  height = 8,
  units = "in",
  device = "png"
)
```

