---
title: "Formatting Rbec output"
author: "Shane Hogle"
date: today
link-citations: true
---

# Introduction

Contains both results from hambiYSKpleiotropy experiment and also a test of the boilprep extraction method from the tipping point experiment.

# Setup

## Libraries
```{r}
#| output: false
library(tidyverse)
library(here)
library(fs)
library(archive)
source(here::here("R", "utils_generic.R"))
```

## Global variables

```{r}
data_raw <- here::here("_data_raw", "communities", "20240318_BTK_illumina_v3")
data <- here::here("data", "communities", "20240318_BTK_illumina_v3")
amplicontar <- here::here(data_raw, "rbec_output.tar.gz")

# make processed data directory if it doesn't exist
fs::dir_create(data)

# create temporary location to decompress
tmpdir <- fs::file_temp()
```

## Data
```{r}
tax_locus_copynum <- tibble::tribble(
     ~strainID, ~rRNA16S_cn, ~rRNA16S_locus,             ~genus,        ~species,
  "HAMBI_0006",          7L,  "H0006_04757",      "Pseudomonas",        "putida",
  "HAMBI_0097",          7L,  "H0097_00044",    "Acinetobacter",     "johnsonii",
  "HAMBI_0097",          7L,  "H0097_02759",    "Acinetobacter",     "johnsonii",
  "HAMBI_0097",          7L,  "H0097_01762",    "Acinetobacter",     "johnsonii",
  "HAMBI_0105",          4L,  "H0105_02306",    "Agrobacterium",   "tumefaciens",
  "HAMBI_0262",          3L,  "H0262_00030",    "Brevundimonas",       "bullata",
  "HAMBI_0403",          9L,  "H0403_00517",        "Comamonas",  "testosteroni",
  "HAMBI_0403",          9L,  "H0403_00522",        "Comamonas",  "testosteroni",
  "HAMBI_1279",          7L,  "H1279_03627",           "Hafnia",         "alvei",
  "HAMBI_1279",          7L,  "H1279_00125",           "Hafnia",         "alvei",
  "HAMBI_1279",          7L,  "H1279_03957",           "Hafnia",         "alvei",
  "HAMBI_1287",          7L,  "H1287_03997",      "Citrobacter",        "koseri",
  "HAMBI_1287",          7L,  "H1287_03402",      "Citrobacter",        "koseri",
  "HAMBI_1292",          7L,  "H1292_03239",       "Morganella",      "morganii",
  "HAMBI_1299",          8L,  "H1299_04293",         "Kluyvera",    "intermedia",
  "HAMBI_1299",          8L,  "H1299_01283",         "Kluyvera",    "intermedia",
  "HAMBI_1299",          8L,  "H1279_03957",         "Kluyvera",    "intermedia",
  "HAMBI_1842",          4L,  "H1842_01650",      "Sphingobium",    "yanoikuyae",
  "HAMBI_1896",          4L,  "H1896_00963", "Sphingobacterium",  "spiritivorum",
  "HAMBI_1972",         10L,  "H1972_00343",        "Aeromonas",        "caviae",
  "HAMBI_1972",         10L,  "H1972_03531",        "Aeromonas",        "caviae",
  "HAMBI_1977",          5L,  "H1977_00118",      "Pseudomonas",  "chlororaphis",
  "HAMBI_1988",          5L,  "H1988_05160",     "Chitinophaga",        "sancti",
  "HAMBI_1988",          5L,  "H1988_05152",     "Chitinophaga",        "sancti",
  "HAMBI_1988",          5L,  "H1988_05165",     "Chitinophaga",        "sancti",
  "HAMBI_2159",          4L,  "H2159_01406",        "Trinickia",   "caryophylli",
  "HAMBI_2159",          4L,  "H2159_05851",        "Trinickia",   "caryophylli",
  "HAMBI_2160",          3L,  "H2160_00530",       "Bordetella",         "avium",
  "HAMBI_2164",          5L,  "H2164_03337",      "Cupriavidus",    "oxalaticus",
  "HAMBI_2443",          3L,  "H2443_00128",       "Paracoccus", "denitrificans",
  "HAMBI_2494",          4L,  "H2494_03389", "Paraburkholderia",   "kururiensis",
  "HAMBI_2659",          4L,  "H2659_00367", "Stenotrophomonas",   "maltophilia",
  "HAMBI_2792",          4L,  "H2792_00549",        "Moraxella",         "canis",
  "HAMBI_3031",          2L,  "H3031_00830",         "Niabella",  "yanshanensis",
  "HAMBI_3237",          6L,  "H3237_00875",       "Microvirga",   "lotononidis",
  "HAMBI_1923",          6L,  "H1923_00876",   "Flavobacterium",      "odoratum"
  )
```

## Functions
```{r}
# this function 
normalize_by_copy <- function(.data, tlc = tax_locus_copynum){
  .data %>% 
    # join with the copy number data frame. We join by the locus tag so this will add H1279_03957 to HAMBI_1299
    dplyr::left_join(tlc, by = join_by(rRNA16S_locus)) %>%
    # get total number of mapping reads per species. This aggregates all the difference ASVs per species
    dplyr::summarize(count = sum(count), .by = c(sample, strainID, rRNA16S_cn)) %>% 
    # group by sample
    dplyr::group_by(sample) %>% 
    # calculate a corrected count which is simply the count divided by copy num for each species
    # dividide by the sum of count divided by copy num for whole sample multiplied by the total
    # number of mapped reads per sample
    dplyr::mutate(count_correct = round(sum(count)*(count/rRNA16S_cn)/sum(count/rRNA16S_cn))) %>%  
    dplyr::ungroup() %>% 
    dplyr::select(sample, strainID, count, count_correct)
  }

# this function replaces missing species counts with zero
completecombos <- function(.data, tlc = tax_locus_copynum, countname = count, remove1923 = TRUE){
 
  # get unique strainIDs
  strainID <- unique(tlc$strainID)
  # table for assigning genus and species names. Doesn't matter if 1923 is there or not
  # because it is filter joined later
  tax <- dplyr::distinct(dplyr::select(tlc, strainID, genus, species))
  if (remove1923) {
    # get unique strainIDs but exclude 1923 if remove1923 is true
    strainID <- strainID[strainID != "HAMBI_1923"]
  }
  
  dplyr::bind_rows(tibble::tibble(strainID = strainID, sample = "dummy"), .data) %>% 
    dplyr::mutate( "{{ countname }}" := dplyr::if_else(sample == "dummy", 1, {{ countname }})) %>% 
    tidyr::complete(sample, strainID) %>% 
    dplyr::filter(sample != "dummy") %>% 
    dplyr::mutate( "{{ countname }}" := dplyr::if_else(is.na({{ countname }}), 0, {{ countname }})) %>% 
    tidyr::replace_na(list(count_correct = 0)) %>% 
    dplyr::left_join(dplyr::distinct(dplyr::select(tlc, strainID, genus, species))) %>% 
    dplyr::relocate(genus, species, .after = strainID)
}
```


# Read metadata

```{r}
mddf <- readr::read_tsv(here::here(data_raw, "20240318_BTK_illumina_v3_metadata.tsv"))
```

# Read Rbec raw counts tables 

## Untar Rbec output tarball

```{r}
archive::archive_extract(
  amplicontar,
  dir = tmpdir,
  files = NULL,
  options = character(),
  strip_components = 0L
)
```

## Setup directory structure

```{r}
tabdir <- here::here(tmpdir, "rbec_output")
samppaths <- fs::dir_ls(tabdir)
sampnames <- fs::path_split(samppaths) %>% 
  purrr::map_chr(dplyr::last)
```

## Read

```{r}
straintabs <- paste0(samppaths, "/strain_table.txt") %>% 
  purrr::set_names(sampnames) %>% 
  purrr::map(
  readr::read_tsv,
  skip = 1,
  col_names = c("rRNA16S_locus","count"),
  show_col_types = FALSE) %>% 
  purrr::list_rbind(names_to = "sample")
```

# Format

Normalize counts by 16S copy number
```{r}
straintabs_norm <- normalize_by_copy(straintabs)
```

Complete all combinations of 23 species
```{r}
straintabs_norm_fmt <- completecombos(straintabs_norm, tlc = tax_locus_copynum, countname = count, remove1923 = TRUE) 
```

```{r}
finaltable <- dplyr::left_join(straintabs_norm_fmt, mddf)
```

```{r}
finaltip <- finaltable %>% 
  filter(project == "tipping_points") %>% 
  dplyr::select(-c(pool:project), -lib_prep, -measure_env, -evolution_env)

finalpleio <- finaltable %>% 
  filter(project == "amanda_chapt01") %>% 
  dplyr::select(-c(pool:project), -lib_prep, -evo_hist, -amp_conc)
```

# Write results

```{r}
readr:::write_tsv(finalpleio, here::here(data, "species_counts_md.tsv"))

readr::write_tsv(finaltip, here::here(data, "species_counts_md_tippingpoints.tsv"))
```

# Clean up

remove decompressed coverage directory from temp location
```{r}
fs::dir_delete(tmpdir)
```
