---
title: "Analysis of growth curves"
author: "Shane Hogle"
date: today
---

# Setup

## Load required packages

```{r}
library(here)
library(tidyverse)
library(stringr)
library(stringi)
library(lubridate)
library(fs)
library(growthrates)
source(here::here("R", "utils_generic.R"))
```

## Global variables

```{r}
#| output: false
#| warning: false
#| error: false
data_raw <- here::here("_data_raw", "monocultures", "20230112_clone_phenotyping")
data <- here::here("data", "monocultures", "20230112_clone_phenotyping")

# make processed data directory if it doesn't exist
fs::dir_create(data)
```

## Functions

```{r}
plotplate <- function(df, dfxy, unsmoothed=TRUE, predicted=FALSE, plate, rows, cols, page){
  dffilt <- dplyr::filter(df, plate == {{ plate }})
  xyfilt <- if (!is.null(dfxy)){ left_join(dfxy, distinct(dffilt, id, well, plate), by = join_by(id)) %>% 
      drop_na()}
  
  ggplot(dffilt, aes(x = hours)) +
    list(
      ggplot2::geom_line(aes(y=OD600_smooth_avg), color = "blue"), 
      if (unsmoothed) {ggplot2::geom_line(aes(y=OD600), color = "orange", lty = 2)},
      if (predicted) {ggplot2::geom_line(aes(y=predicted), color = "orange")}, 
      if (!is.null(dfxy)) {ggplot2::geom_point(data = xyfilt, aes(x = x, y = y), color = "red", size = 2)},
      ggplot2::labs(x = "Hours", y = "OD600"), 
      ggplot2::scale_x_continuous(breaks = seq(0, 48, 12), labels = seq(0, 48, 12)), 
      ggforce::facet_wrap_paginate(~ well, nrow = rows, ncol = cols, page = page), #scales = "free_y"
      ggplot2::theme(axis.text = element_text(size = 9))
    )
}
```

# Read data

## Metadata

```{r}
mddf <- read_tsv(
  here::here(data_raw, "metadata.tsv"),
  col_types = list(
    plate = col_character(),
    well = col_character(),
    clone_id = col_double(),
    experiment_treatment = col_character(),
    LTCSE_treatment = col_character(),
    replicate = col_character(),
    lagtime = col_time(format = "%T"),
    max_growth_death_rate = col_double(),
    time_to_stationary = col_time(format = "%T")
  )
) %>% mutate(Group = 1, Flag = 0, Subset = 1)
```

## Previously smoothed growth curves

```{r}
gcurves_thin_sm <- readr::read_tsv(here::here(data, "gcurves_smoothed_thinned.tsv"))
```


# Spline based estimate for growth rate

Using the tool [`growthrates`](https://cran.r-project.org/web/packages/growthrates/index.html) to estimate mu_max. I have found this works a lot better the gcplyr and is more convenient than using another tool outside of R. Nonparametric estimate growth rates by spline is very fast. Fitting to a model takes more time resources. [Generally it is best to try multiple approaches and to visualize/check the data to make sure it makes sense.](https://www.frontiersin.org/journals/ecology-and-evolution/articles/10.3389/fevo.2023.1313500/full)

```{r}
gcurves_thin_sm <- gcurves_thin_sm %>% 
  # make uniq id
  mutate(id = paste0(plate, "|", well))
```

Smoothing splines are a quick method to estimate maximum growth. The method is called nonparametric, because the growth rate is directly estimated from the smoothed data without being restricted to a specific model formula.

From [growthrates documentation:](https://cran.r-project.org/web/packages/growthrates/growthrates.pdf)

> The method was inspired by an algorithm of [Kahm et al. (2010)](https://www.jstatsoft.org/article/view/v033i07), with different settings and assumptions. In the moment, spline fitting is always done with log-transformed data, assuming exponential growth at the time point of the maximum of the first derivative of the spline fit. All the hard work is done by function smooth.spline from package stats, that is highly user configurable. Normally, smoothness is automatically determined via cross-validation. This works well in many cases, whereas manual adjustment is required otherwise, e.g. by setting spar to a fixed value \[0, 1\] that also disables cross-validation.

### Fit

```{r}
#| eval: false
set.seed(45278)
many_spline <- growthrates::all_splines(OD600_smooth_avg ~ hours | id, data = gcurves_thin_sm, spar = 0.5)

readr::write_rds(many_spline, here::here(data, "spline_fits"))
```

```{r}
#| echo: false
#| warning: false
many_spline <- readr::read_rds(here::here(data, "spline_fits"))
```

### Results

```{r}
many_spline_res <- growthrates::results(many_spline)
```

### Write output
```{r}
many_spline_res %>% 
  dplyr::left_join(dplyr::distinct(dplyr::select(gcurves_thin_sm, id, plate, well)), by = join_by(id)) %>% 
  dplyr::select(-id) %>% 
  dplyr::relocate(y0:r2, .after="well") %>% 
  readr::write_tsv(here::here(data, "gcurve_spline_results.tsv"))
```

### Predictions

```{r}
many_spline_xy <- purrr::map(many_spline@fits, \(x) data.frame(x = x@xy[1], y = x@xy[2])) %>%
  purrr::list_rbind(names_to = "id")

many_spline_fitted <- purrr::map(many_spline@fits, \(x) data.frame(x@FUN(x@obs$time, x@par))) %>%
  purrr::list_rbind(names_to = "id") %>%
  dplyr::rename(hours = time, predicted = y) %>%
  dplyr::left_join(gcurves_thin_sm, by = dplyr::join_by(id, hours)) %>%
  dplyr::group_by(id) %>%
  # this step makes sure we don't plot fits that go outside the range of the data
  dplyr::mutate(predicted = dplyr::if_else(dplyr::between(
    predicted, min(OD600_smooth_avg), max(OD600_smooth_avg * 1.5)
  ), predicted, NA_real_)) %>%
  dplyr::ungroup()
```

### Plot

# Inspect growth curves

The subscripts at the ends of the plate contain informations:

1. The `_1` subscript is the "bact" condition with no additional treatment
2. The `_2` subscript is the "bact_pred" condition which is a ciliate predator treatment
3. The `_3` subscript is the "bact_strep" condition which has streptomycin
4. The `_4` subscript is the "bact_pred_strep" condition which has both streptomycin and ciliates present

The same well always contains the same clone such that well "A1" on plate A1 will be the same clone phenotyped just as itself, with a predator, with streptomycin, and with both a predator and streptomycin

## Bact conditions

### Plate 1

::: {#fig-01}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A1_1", rows=8, cols=12, page=1)
```
Growth curves for the **bacteria only** conditions on **Plate 1**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 2

::: {#fig-02}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A2_1", rows=8, cols=12, page=1)
```
Growth curves for the **bacteria only** conditions on **Plate 2**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 3

::: {#fig-03}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A3_1", rows=8, cols=12, page=1)
```
Growth curves for the **bacteria only** conditions on **Plate 3**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 4

::: {#fig-04}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A4_1", rows=8, cols=12, page=1)
```
Growth curves for the **bacteria only** conditions on **Plate 4**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 5

::: {#fig-05}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A5_1", rows=8, cols=12, page=1)
```
Growth curves for **bacteria only** conditions on **Plate 5**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 6

::: {#fig-06}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A6_1", rows=8, cols=12, page=1)
```
Growth curves for **bacteria only** conditions on **Plate 6**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 7

::: {#fig-07}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A7_1", rows=8, cols=12, page=1)
```
Growth curves for **bacteria only** conditions on **Plate 7**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 8

::: {#fig-08}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A8_1", rows=8, cols=12, page=1)
```
Growth curves for **bacteria only** conditions on **Plate 8**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

## Predator conditions

### Plate 1

::: {#fig-09}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A1_2", rows=8, cols=12, page=1)
```
Growth curves for the **ciliate** conditions on **Plate 1**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 2

::: {#fig-10}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A2_2", rows=8, cols=12, page=1)
```
Growth curves for the **ciliate** conditions on **Plate 2**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 3

::: {#fig-11}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A3_2", rows=8, cols=12, page=1)
```
Growth curves for the **ciliate** conditions on **Plate 3**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 4

::: {#fig-12}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A4_2", rows=8, cols=12, page=1)
```
Growth curves for the **ciliate** conditions on **Plate 4**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 5

::: {#fig-13}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A5_2", rows=8, cols=12, page=1)
```
Growth curves for **ciliate** conditions on **Plate 5**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 6

::: {#fig-14}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A6_2", rows=8, cols=12, page=1)
```
Growth curves for **ciliate** conditions on **Plate 6**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 7

::: {#fig-15}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A7_2", rows=8, cols=12, page=1)
```
Growth curves for **ciliate** conditions on **Plate 7**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 8

::: {#fig-16}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A8_2", rows=8, cols=12, page=1)
```
Growth curves for **ciliate** conditions on **Plate 8**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

## Streptomycin conditions

### Plate 1

::: {#fig-17}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A1_3", rows=8, cols=12, page=1)
```
Growth curves for the **Streptomycin** conditions on **Plate 1**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 2

::: {#fig-18}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A2_3", rows=8, cols=12, page=1)
```
Growth curves for the **Streptomycin** conditions on **Plate 2**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 3

::: {#fig-19}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A3_3", rows=8, cols=12, page=1)
```
Growth curves for the **Streptomycin** conditions on **Plate 3**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 4

::: {#fig-20}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A4_3", rows=8, cols=12, page=1)
```
Growth curves for the **Streptomycin** conditions on **Plate 4**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 5

::: {#fig-21}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A5_3", rows=8, cols=12, page=1)
```
Growth curves for **Streptomycin** conditions on **Plate 5**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 6

::: {#fig-22}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A6_3", rows=8, cols=12, page=1)
```
Growth curves for **Streptomycin** conditions on **Plate 6**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 7

::: {#fig-23}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A7_3", rows=8, cols=12, page=1)
```
Growth curves for **Streptomycin** conditions on **Plate 7**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 8

::: {#fig-24}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A8_3", rows=8, cols=12, page=1)
```
Growth curves for **Streptomycin** conditions on **Plate 8**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

## Predator + Streptomycin conditions

### Plate 1

::: {#fig-25}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A1_4", rows=8, cols=12, page=1)
```
Growth curves for the **Ciliate + Streptomycin** conditions on **Plate 1**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 2

::: {#fig-26}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A2_4", rows=8, cols=12, page=1)
```
Growth curves for the **Ciliate + Streptomycin** conditions on **Plate 2**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 3

::: {#fig-27}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A3_4", rows=8, cols=12, page=1)
```
Growth curves for the **Ciliate + Streptomycin** conditions on **Plate 3**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 4

::: {#fig-28}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A4_4", rows=8, cols=12, page=1)
```
Growth curves for the **Ciliate + Streptomycin** conditions on **Plate 4**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 5

::: {#fig-29}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A5_4", rows=8, cols=12, page=1)
```
Growth curves for **Ciliate + Streptomycin** conditions on **Plate 5**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 6

::: {#fig-30}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A6_4", rows=8, cols=12, page=1)
```
Growth curves for **Ciliate + Streptomycin** conditions on **Plate 6**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 7

::: {#fig-31}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A7_4", rows=8, cols=12, page=1)
```
Growth curves for **Ciliate + Streptomycin** conditions on **Plate 7**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::

### Plate 8

::: {#fig-32}
```{r}
#| fig.width: 12
#| fig.height: 8
#| echo: false
#| warning: false
plotplate(many_spline_fitted, dfxy=many_spline_xy, unsmoothed=FALSE, predicted=TRUE, plate="A8_4", rows=8, cols=12, page=1)
```
Growth curves for **Ciliate + Streptomycin** conditions on **Plate 8**. Blue line is smoothed with a moving average window of 5 points. Orange is slope of max predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
:::


