---
title: "Jensen-Shannon divergence of communities"
subtitle: "Community composition workflow"
author: "Shane Hogle"
date: today
abstract: "This notebook compares the response of different YSK evolutionary histories to the ancestral community in each experimental measurement environment using pairwise Jensen-Shannon divergences."
---

# Setup

Loads required libraries and sets global variables

```{r}
#| output: false
library(tidyverse)
library(here)
library(fs)
library(scales)
library(philentropy)
source(here::here("R", "utils_generic.R"))
```

# Read

Read species abundance data (16S v3 amplicon counts) and do some light formatting of metadata 

```{r}
#| output: false
#| warning: false
sptable <- readr::read_tsv(here::here(data_sp, "species_counts_md.tsv")) %>% 
  dplyr::mutate(transfer = day/7)

counts_f <- sptable %>% 
  dplyr::group_by(sample) %>% 
  dplyr::mutate(f=count_correct/sum(count_correct)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(measure_env_short = dplyr::case_when(measure_env == "bact" ~ "Meas: B",
                                       measure_env == "bact_strep" ~ "Meas: BS")) %>% 
  dplyr::mutate(home_env_short = dplyr::case_when(
                                    evolution_env == "anc" ~ "Home: Anc",
                                    evolution_env == "bact" ~ "Home: B",
                                    evolution_env == "bact_strep" ~ "Home: BS")) %>% 
  dplyr::mutate(
    measure_env_short = factor(measure_env_short, levels = c("Meas: B", "Meas: BS")),
    home_env_short = factor(home_env_short, levels = c("Home: Anc", "Home: B", "Home: BS")),
         day = factor(day),
         replicate = factor(replicate),
         strainID = factor(strainID, levels = names(hambi_colors)))
```

# Formatting

Some light formatting to subset data into distinct tibbles for later plotting

```{r}
#| output: false
#| warning: false
# these are communities of a (supposedly) known composition. Can be used with metacal
pos_ctrl_samples <- counts_f %>% 
  dplyr::filter(str_detect(sample, "pos_ctrl"))

# these are samples taken directly from YSK and represent the composition of the communities used to start the experiment
t0_samples <- counts_f %>% 
  dplyr::filter(!str_detect(sample, "pos_ctrl")) %>% 
  dplyr::filter(day == 0)

# only samples from the experiment
counts_f_experiment <- dplyr::anti_join(counts_f, pos_ctrl_samples) %>% 
  dplyr::anti_join(., t0_samples) %>% 
  dplyr::mutate(
    measure_env_short = factor(measure_env_short, levels = c("Meas: B", "Meas: BS")),
    home_env_short = factor(home_env_short, levels = c("Home: Anc", "Home: B", "Home: BS")),
         day = factor(day),
         replicate = factor(replicate),
         strainID = factor(strainID, levels = names(hambi_colors)))
```


# Distances between communities

The Jensen–Shannon divergence (JSD) is a symmetrized and smoothed version of the [Kullback–Leibler divergence](https://en.wikipedia.org/wiki/Kullback%E2%80%93Leibler_divergence "Kullback–Leibler divergence"). It is defined as:

$$
D_{\text{JS}}(P \parallel Q) = \frac{1}{2} D_{\text{KL}}(P \parallel M) + \frac{1}{2} D_{\text{KL}}(Q \parallel M)
$$ {#eq-JSD}

$$
M = \frac{1}{2}(P + Q)
$$ {#eq-avg_dist}

Here, ($D_{\text{KL}}$) denotes the Kullback-Leibler divergence which is defined as:

$$
D_{KL}(P \parallel Q) = \sum_{x \in \mathcal{X}} P(x) \log \frac{P(x)}{Q(x)}
$$ {#eq-KLD}

## JSD to ancestor no samples filtered

```{r}
#| output: false
#| warning: false
# create a matrix of species frequencies
prob_mat <- counts_f %>% 
  dplyr::select(sample, strainID, f) %>% 
  tidyr::pivot_wider(names_from = strainID, values_from = f) %>% 
  tibble::column_to_rownames(var = "sample") %>% 
  as.matrix()

# calculate JSD between all pairs using philentropy::JSD
jsd_mat <- philentropy::JSD(prob_mat)

# get nonredundant pairings
ind <- which(upper.tri(jsd_mat, diag = TRUE), arr.ind = TRUE)
nn <- list(rownames(prob_mat), rownames(prob_mat))
jsddf <- data.frame(sample.x = nn[[1]][ind[, 1]],
           sample.y = nn[[2]][ind[, 2]],
           jsd = jsd_mat[ind])

# metadata for joining
md <- dplyr::distinct(counts_f, sample, replicate, transfer, measure_env_short, home_env_short)

# convert to long format
jsddf_long_md <- jsddf %>% 
  dplyr::left_join(md, by = join_by(sample.x==sample)) %>% 
  dplyr::left_join(md, by = join_by(sample.y==sample))
```

Filter JSD pairs to only those that are relative to ancestral community in each measurement environment. Also calculate mean and boostrapped 95CI using HMisc

```{r}
#| output: false
#| warning: false
jsddf_long_md_filt <- jsddf_long_md %>% 
  dplyr::filter(measure_env_short.x == measure_env_short.y) %>% 
  dplyr::filter(home_env_short.x != home_env_short.y & transfer.x == transfer.y & home_env_short.x == "Home: Anc") %>% 
  dplyr::summarize(ggplot2::mean_cl_boot(jsd), .by=c(transfer.y, home_env_short.y, measure_env_short.y)) %>% 
  dplyr::mutate(home_env_short.y = factor(home_env_short.y))
```

Hmmm.... not sure what exactly this is telling, but I am calculating the within home/measurement divergence as some kind of variability baseline.

```{r}
intrarepvar <- jsddf_long_md %>% 
  dplyr::filter(measure_env_short.x == measure_env_short.y & home_env_short.x == home_env_short.y & transfer.x == transfer.y) %>% 
  dplyr::summarize(ggplot2::mean_cl_boot(jsd), .by=c(transfer.y, home_env_short.y, measure_env_short.y)) %>% 
  dplyr::filter(home_env_short.y != "Home: Anc") %>% 
  dplyr::mutate(home_env_short.y = factor(home_env_short.y))
```

Make plot of JSD to ancestral communities in each condition and save in raster/vector format

```{r}
plot_jsd <- function(intra_df, df){
  mycols <- c("Home: B" = "#66c2a5", "Home: BS" = "#fc8d62")
  ggplot2::ggplot() +
  ggplot2::geom_ribbon(data = intra_df, 
              aes(x = transfer.y, ymin = ymin, ymax=ymax, fill = home_env_short.y), alpha=0.25) + 
  ggplot2::geom_line(data = intra_df, 
            aes(x = transfer.y, y=y, color = home_env_short.y), alpha=0.5, lty=2) + 
  ggplot2::geom_point(data = df, aes(x = transfer.y, y = y, group = home_env_short.y, color = home_env_short.y)) +
  ggplot2::geom_linerange(data = df, 
                 aes(x = transfer.y, ymin = ymin, ymax = ymax, color = home_env_short.y)) + 
  ggplot2::geom_line(data= df, 
            aes(x = transfer.y, y = y, group = home_env_short.y, color = home_env_short.y)) + 
  ggplot2::facet_grid(measure_env_short.y ~ home_env_short.y, scales = "free_y") + 
  ggplot2::labs(x = "Growth Cycles", y = "Jensen-Shannon Divergence relative to ancestor", color = "Home\nEnvironment", fill = "Home\nEnvironment") +
  #ggplot2::scale_y_sqrt() +
  ggplot2::scale_x_continuous(breaks = c(4, 8, 12), limits = c(3.5, 12.5)) +
  ggplot2::scale_color_manual(values=mycols) +
  ggplot2::scale_fill_manual(values=mycols) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    #panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    strip.placement = 'outside',
    strip.background = element_blank(),
  )
}

pjsd <- plot_jsd(intrarepvar, jsddf_long_md_filt)
```

```{r}
#| warning: false
#| eval: false
ggplot2::ggsave(
  here::here(figs, "community_jsd_to_anc.svg"),
  pjsd,
  width = 8,
  height = 6,
  units = "in",
  device = "svg"
)
```

::: {#fig-03}
```{r}
#| fig.width: 8
#| fig.height: 6
#| echo: false
#| warning: false
pjsd
```
Mean $\pm$ 95% CI of the pairwise Jensen-Shannon Divergence between the ancestral/clonal community composition (@fig-01 C) and the community composition in YSK Home environments (Home, grid columns) for each measurement environment (Meas, grid rows) in the transplantation experiment (see also @fig-01 D). Mean divergences are nested within Growth Cycles (horizontal axis) where one growth cycle is 7 days. Dashed lines indicate mean and ribbons indicate 95% CI for the mean within-condition Jensen-Shannon Divergence and is meant as an indicator of replicate variability within each Home/Measurement treatment combination.
:::

## JSD to ancestor - idiosyncratic samples filtered 

In @fig-01 is it clear that some of the replicates in samples from YSK have a very different composition from the others. This is likely due to rapid resistance evolution of HAMBI 1972 in those particular replicates from YSK (will need to verify with the metagenomes). HAMBI_1972 is normally very sensitive to STR. The idea in this section is to repeat the JSD to ancestor analysis but exclude these idiosyncratic replicates - A and F from YSK BS, C and E from YSK B. We want to make sure that the trends in JSD relative to the ancestral community is not being driven primarily by these outlier replicates.

Filter JSD pairs to only those that are relative to ancestral community in each measurement environment. Also exclude the replicates mentioned above. Finally calculate mean and boostrapped 95CI using HMisc.

```{r}
#| output: false
#| warning: false
jsddf_long_md_filt2 <- jsddf_long_md %>% 
  dplyr::filter(measure_env_short.x == measure_env_short.y) %>% 
  dplyr::filter(home_env_short.x != home_env_short.y & transfer.x == transfer.y & home_env_short.x == "Home: Anc") %>% 
  dplyr::filter(!((measure_env_short.x == "Meas: BS" & home_env_short.y == "Home: BS") & str_detect(replicate.x, "A|F") | str_detect(replicate.y, "A|F"))) %>% 
  dplyr::filter(!((measure_env_short.x == "Meas: BS" & home_env_short.y == "Home: B") & str_detect(replicate.x, "C|E") | str_detect(replicate.y, "C|E"))) %>% 
  dplyr::summarize(ggplot2::mean_cl_boot(jsd), .by=c(transfer.y, home_env_short.y, measure_env_short.y)) %>% 
  dplyr::mutate(home_env_short.y = factor(home_env_short.y))
  
```

Calculating the within home/measurement divergence as some kind of variability baseline but with the outlier replicates excluded.

```{r}
intrarepvar2 <- jsddf_long_md %>% 
  dplyr::filter(measure_env_short.x == measure_env_short.y & home_env_short.x == home_env_short.y & transfer.x == transfer.y) %>% 
    dplyr::filter(!((measure_env_short.x == "Meas: BS" & home_env_short.y == "Home: BS") & str_detect(replicate.x, "A|F") | str_detect(replicate.y, "A|F"))) %>% 
  dplyr::filter(!((measure_env_short.x == "Meas: BS" & home_env_short.y == "Home: B") & str_detect(replicate.x, "C|E") | str_detect(replicate.y, "C|E"))) %>% 
  dplyr::summarize(ggplot2::mean_cl_boot(jsd), .by=c(transfer.y, home_env_short.y, measure_env_short.y)) %>% 
  dplyr::filter(home_env_short.y != "Home: Anc") %>% 
  dplyr::mutate(home_env_short.y = factor(home_env_short.y))
```

Make plot of JSD to ancestral communities in each condition and save in raster/vector format

```{r}
pjsd2 <- plot_jsd(intrarepvar2, jsddf_long_md_filt2)
```

::: {#fig-04}
```{r}
#| fig.width: 8
#| fig.height: 6
#| echo: false
#| warning: false
pjsd2
```
Same as in @fig-03 but with excluding the idiosyncratic replicates A and F from YSK Strep and C and E from YSK Bact.
:::

So here it does appear that the trends still hold but now the JSD to ancestor always exceeds variation within each environment/evolutionary history.

## JSD to YSK founder community

Here I just wanted to check how each community with an YSK evolutionary history compares to the original YSK source material

::: {#fig-05}
```{r}
#| fig.width: 8
#| fig.height: 6
#| warning: false
jsddf_long_md %>% 
  dplyr::filter(transfer.x == 0 | transfer.y == 0) %>% 
  dplyr::filter(home_env_short.x != "Home: Anc") %>% 
  dplyr::filter(str_remove(measure_env_short.y, "Meas: ") == str_remove(home_env_short.y, "Home: ")) %>% 
  #dplyr::filter(home_env_short.x == home_env_short.y) %>% 
  dplyr::filter(replicate.x == replicate.y) %>% 
  ggplot(aes(x= replicate.x, y = jsd)) +
  geom_jitter(aes(color = factor(transfer.y)), height = 0.01, width = 0.1) + 
  facet_grid(measure_env_short.y ~ home_env_short.x) +
  ggplot2::labs(x = "Replicate microcosm", y = "Jensen-Shannon Divergence relative YSK source", color = "Transfer") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    strip.placement = 'outside',
    strip.background = element_blank(),
  )

```
JSD relative to 58 week YSK source material (nested by replicate) for each treatment/YSK history and transfer in the main experiment. Excludes sames that were inoculated with the ancestral community.
:::